Semantic Linearity in a Lazy Optimising Compiler

### 2025-10-07 popl26-accepted-before-rebuttal.tex

%%% \subsection{Consuming tagged resources as needed}
%%% 
%%% As discussed in Section~\ref{}, pattern-bound linear variables are
%%% put in the context with a \emph{tagged} usage environment with the resources of
%%% the scrutinee. In a \emph{tagged} usage environment environment, all resources
%%% are tagged with a constructor and an index into the many fields of the
%%% constructor.
%%% 
%%% In practice, a resource might have more than one tag. For example, in the following
%%% program, after the first pattern match, |a| and |b| have, respectively, usage
%%% environments $\{\lctag{x}{K_1}\}$ and $\{\lctag{x}{K_2}\}$:
%%% \begin{code}
%%% f x = case x of
%%%        K a b -> case a of
%%%         Pair n p -> (n,p)
%%% \end{code}
%%% However, in the following alternative, |n| has usage environment
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_1}\}$ and |p| has
%%% $\{\lctag{\lctag{x}{K_1}}{Pair_2}\}$. To typecheck
%%% |(n,p)|, one has to $Split$ |x| first on |K| and then on |Pair|, in order for
%%% the usage environments to match.
%%% 
%%% In our implementation, we split resources on demand (and don't directly allow
%%% splitting linear resources), i.e. when we use a tagged resource we split the
%%% linear resource in the linear environment (if available), but never split otherwise.
%%% %
%%% Namely, starting on the innermost tag (the closest to the variable name), we
%%% substitute the linear resource for its split fragments, and then we iteratively
%%% further split those fragments if there are additional tags.
%%% %
%%% We note that it is safe to destructively split the resource (i.e. removing the
%%% original and only leaving the split fragments) because we only split resources
%%% when we need to consume a fragment, and as soon as one fragment is consumed
%%% then using the original ``whole'' variable would violate linearity.
%%% 
%%% In the example, if |n| is used, we have to use its usage environment, which in
%%% turn entails using $\lctag{\lctag{x}{K_1}}{Pair_1}$, which has two tags. In this order, we:
%%% \begin{itemize}
%%% \item Split $x$ into $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$
%%% \item Split $\lctag{x}{K_1}$ and $\lctag{x}{K_2}$ into
%%%   \begin{itemize}
%%%   \item $\lctag{\lctag{x}{K_1}}{Pair_1}$ and $\lctag{\lctag{x}{K_1}}{Pair_2}$
%%%   \item Leave $\lctag{x}{K_2}$ untouched, as we only split on demand, and we aren't using a fragment of $\lctag{x}{K_2}$.
%%%   \end{itemize}
%%% \item Consume $\lctag{\lctag{x}{K_1}}{Pair_1}$, the usage environment of $n$, by removing it from the typing environment.
%%% \end{itemize}

%%%\subsection{Merging Linear Core into GHC\label{sec:merging-linear-core}}
%%%
%%%Describe the ticket for linear Core, the pending MRs, and the difficulty in
%%%even annotating the bind site across optimisations regardless of multiplicities.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related and Future Work\label{sec:discussion}}

% In this chapter, we compare our contributions and Linear Core to related
% existing works in the literature,
% % (including linearity in a lazy, evaluation models in terms of linearity, and
% % the Core system).
% %
% consider further research
% % deemed out of the scope of this work and of the Linear Core type system
% (notably, so-called \emph{multiplicity coercions} to handle the
% interaction between linearity and \emph{coercions}, a key feature of Core which we
% left out our system), and conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Related Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%paragraph{Related Work}

% In this section we discuss related work, namely, Linear
% Haskell~\cite{cite:linearhaskell}, Linear Mini-Core~\cite{cite:minicore}, and
% linearity-influenced optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell}.

% TODO: A brief introduction to the related work section?

% \subsection{Formalization of Core}\todo{terrible paragraph name, and terrible paragraph}
% 
% As such, there exists no formal definition of Core that
% accounts for linearity. In this context, we intend to introduce a linearly typed
% System $F_C$ with multiplicity annotations and typing rules to serve
% as a basis for a linear Core. Critically, this Core linear language
% must account for call-by-need evaluation semantics and be valid in
% light of Core-to-Core optimising transformations.

% \parawith{System FC}

%System $F_C$~\cite{cite:systemfc} (Section~\ref{sec:core}) does not
%account for linearity in its original design, and, to the best of our
%knowledge, no extension to System $F_C$ with linearity and non-strict
%semantics exists.
%

% \begin{itemize}
% \item SystemFC tal como está não tem linearidade de todo
% \item Formalmente nao temos published definição de linearidade no Core
% \item Regras para sistema tipo FC com linearidade
% \item mas uma extensão tipo linear lambda calculus nao consegui exprimir as transformações do core
% \end{itemize}

% Haskell Core's foundational language was imbued with linear types, but it does
% not account for linearity with the whole of the type system
% 
% Multiplicity annotations in SystemFC?
% 
% Rules?

\paragraph{Linear Haskell\label{sec:related-work-linear-haskell}.}
%
% \todo[inline]{Say something about Linear Haskell's lazy operational semantics,
% but the type system not being concerned with linearity in the presence of
% laziness}
%
% Haskell, contrary to most programming languages with linear types, has existed
% for 31 years of its life \emph{without} linear types. As such, the introduction
% of linear types to Haskell comes with added challenges that do not exist in
% languages that were designed with linear types from the start:
% %
% \begin{itemize}
%     \item Backwards compatibility. The addition of linear types shouldn't break
%         all existing Haskell code.
%     \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
%         its non-linearly-typed counterpart should fit in together and it must be
%         possible to define functions readily usable by both sides
%         simultaneously.
%     \item Future-proofing. Haskell, despite being an
%         industrial-strength language, is also a petri-dish for experimentation
%         and innovation in the field of programming languages. Therefore, Linear
%         Haskell takes care to accommodate possible future features, in
%         particular, its design is forwards compatible with affine and dependent
%         types.
% \end{itemize}
% %
Linear Haskell~\cite{cite:linearhaskell} augments the Haskell surface language
with linear types, but it is not concerned with extending GHC's intermediate
language(s), which presents its own challenges. 
%
However, the implementation of Linear Haskell in GHC does modify and extend Core
with linearity/multiplicity annotations. Core's type system is unable to type
semantic linearity of programs in the sense elaborated in our work,
which results in Core rejecting most linear programs resulting from optimising
transformations that leverage the non-strict semantics of Core.
%
Our work overcomes the limitations of Core's linear type system derived from
Linear Haskell by understanding linearity semantically in the presence of
laziness, and by showing that multiple Core-to-Core
optimisations employed by GHC are linearity preserving, using a
linearity-aware semantics that is essentially identical to that of~\cite{cite:linearhaskell}.
Linear Core can also be seen as a system that validates the programs written in
Linear Haskell and are compiled by GHC, by guaranteeing (through typing) that
linear resources are still used exactly once throughout the optimising
transformations.

\paragraph{Linearity-aware Semantics}
Several other works explore linearity aware semantics,
dating back to \cite{Chirimar_Gunter_Riecke_1996} and
\cite{TURNER1999231}.
The former uses a computational interpretation of linear logic to give
an account of reference counting. The latter present a heap-based
operational interpretation of linear logic and explore memory usage
properties under call-by-name and call-by-need, showing both satisfy
different properties under this lens. Their semantics is similar to
ours, but does not account for pattern matching.
More recently, \cite{DBLP:journals/pacmpl/ChoudhuryEEW21,DBLP:conf/esop/MarshallVO22} and
\cite{cite:linearhaskell} all make use of usage-aware
semantics. \citet{DBLP:journals/pacmpl/ChoudhuryEEW21} do so in the context of graded dependent
type theory, using the usage-aware heap-based operational semantics to
show correct accounting of resource usage in their setting, as we do
in ours, but they do not model
call-by-need, only call-by-name. The work of \cite{DBLP:conf/esop/MarshallVO22} is
applies the ideas of \cite{DBLP:journals/pacmpl/ChoudhuryEEW21}
to a non-graded (also call-by-name) setting. \citet{cite:linearhaskell} provide a similar semantics
to ours, but present only a naive linear typing system that does not
leverage non-strict evaluation.

\paragraph{Linear Mini-Core\label{sec:linear-mini-core}}

Linear Mini-Core~\cite{cite:minicore} is a specification of linear types in
Core as they were being implemented in GHC, and doubles as the (unpublished)
precursor to our work. Linear Mini-Core first observes the incapacity of
Core's type system to accept linear programs after transformations, and 
introduces usage environments for let-bound variables with the same goal of
Linear Core of specifying a linear type system for Core that accepts the
optimising transformations.

We draw from Linear Mini-Core the rule for non-recursive let expressions and
how let-bound variables are annotated with a usage environment. However, our
work further explores the interaction of laziness with linearity in depth, and
diverges significantly in rules for typing other constructs, notably, case expressions and
case alternatives. Furthermore, unlike Mini-Core, we prove Linear Core is type
safe, guarantees linear resource usage, and that multiple optimising
transformations, when applied to Linear Core programs, preserve linearity.

% \subsection{OutsideIn(X)\label{related-work-gadts}}
% 
% Defines constraint-based type system parametrized over X which does not account
% for local type refinements regarding linearity.
% 
% Se for modificar o typechecker com as multiplicity coercions vou ter de falar
% disto.

% \subsection{Rust}
% 
% Rust has a core based on linear types. Describe Rust's architecture?
% How do they handle linearity plus optimisations
% They probabluy don't typecheck linearity in Core

\paragraph{Linearity-influenced optimisations}

Core-to-Core transformations appear in many works across the research
literature~\cite{cite:let-floating,peytonjones1997a,santos1995compilation,peytonjones2002secrets,baker-finch2004constructed,maurer2017compiling,Breitner2016_1000054251,sergey_vytiniotis_jones_breitner_2017},
all designed in the context of a typed language (Core) which does not have
linear types. However, some works~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell} have observed that
certain optimisations (in particular, let-floating and inlining) greatly
benefit from linearity analysis and, in order to improve those transformation,
special purpose linear-type-inspired systems were created and implemented.
%
Preserving linear types in Core throughout the compilation pipeline allows the
optimiser to leverage non-heuristic linearity information where, previously, it
would resort solely to ad-hoc or incomplete custom-built linearity inference
passes (naturally, these passes are still necessary to optimise programs not
using explicit linearity or multiplicity annotations). Linearity may
potentially be used to the benefit of other optimising transformations that
currently do not take it into account.

% \begin{itemize}
% \item Transfs. core to core aparecem em vários artigos, e são desenhadas no contexto de uma linguagem tipificada mas que não é linearly typed.
% \item nestes dois artigos é observado que se houvesse a capacidade de explorar linearidade podiamos fazer as coisas de forma diferente
% \item Todas estas optimizaçoes de decadas foram desenhadas sem linear types e há sitios onde linear types podiam ajudar mas não existiam na altura
% \item POdemos usar linear types multiplicitpiadads para lazy language core q definimos para nao ter de fazer sistemas lineares de proposito para optimizações
% \item Ser ad-hoc incompleto ou nao feito de todo
% \end{itemize}

% \parawith{A transformation based optimiser for Haskell}
% They discuss a cardinality analysis based on a linear type system but create (an
% ad-hoc?) one suited. Comparison in the measure of creating optimizations based
% on linearity.
% 
% \parawith{Let-Floating: Moving Bindings to Give Faster Programs\label{sec:rw:let-floating}}
% In the paper~\cite{cite:let-floating}...
% They say they are doing work on a linear type system to identify places where
% the lambda is linearly used and therefore floating-in is beneficial and
% floating-out not as productive.

% \subsection{Call-by-value, call-by-name and call-by-value...}
% 
% The work~\cite{cite:call-by-name-value-and-need-linear-lambda-calculus}...

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Future Work
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Future Work\label{sec:future-work}}

We highlight some avenues of future work. Briefly,
these include \emph{multiplicity coercions}, optimisations leveraging
linearity, resource inference for usage environments, and ultimately using
Linear Core in a mature optimising compiler with lazy evaluation and linear
types -- the Glasgow Haskell Compiler. Lastly, we discuss the
% we hypothesize Linear($X$), a linear system parametrized on the evaluation strategy, and the
generalization of Linear Core to the surface Haskell language.

\paragraph{Multiplicity Coercions.}
Linear Core does not have type equality coercions, a flagship feature of GHC
Core's type system.
%
Coercions %briefly explained in Section~\ref{sec:core},
allow the Core intermediate language to encode a panoply of Haskell source
type-level features such as GADTs, type families, or newtypes.
%
In Linear Haskell, multiplicities are introduced as annotations to function
arrows which specify the linearity of the function. In practice,
multiplicities are simply types of kind \incode{Multiplicity}, where \incode{One} and \incode{Many}
are the type constructors of the kind \incode{Multiplicity}; multiplicity polymorphism
follows from type polymorphism, where multiplicity variables are
just type variables. Encoding multiplicities as types allows Haskell programs
to leverage features available for types to naturally extend to multiplicities
as well.
For instance, it is possible via the use of GADTs to define a
function whose linearity of its second argument depends on the value
of its first argument, internally realized through so-called
\emph{multiplicity coercions}. Currently, Core cannot make use of such
coercions to determine whether the usages of linear resources match
their intended multiplicity. Studying the interaction between coercions and multiplicities is a main avenue
of future work for Linear Core.
%
% Consequently, we might define, e.g., using a GADT \incode{SBool} and a type family
% |If|, the function \incode{dep} which is linear in the second argument if the first
% argument is \incode{STrue} and unrestricted otherwise:
% %
% \begin{limitation}
% \begin{lstlisting}
% data SBool :: Bool -> Type where
%   STrue :: SBool True
%   SFalse :: SBool False

% type family If b t f where
%   If True t _ = t
%   If False _ f = f

% dep :: SBool b -> Int %(If b One Many) -> Int
% dep STrue x = x
% dep SFalse _ = 0
% \end{lstlisting}
% \end{limitation}
% %
% This example is linear and should be accepted. However, the example is rejected
% by the GHC's Core type checker.
% Critically, Core doesn't currently understand
% so-called \emph{multiplicity coercions}. Even though after matching on |STrue|
% we have access to a coercion from the function multiplicity $m$ to $1$ ($m \sim
% 1$), we cannot use this coercion to determine whether the usages of the linear
% resources match the multiplicity.
% %
% Studying the interaction between coercions and multiplicities is a main avenue
% of future work for Linear Core.

% In GHC, multiplicity coercions are tracked by issue $19517$\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/19517}.

% \todo[inline]{Relate to levity polymorphism and runtime rep coercions (Sam)}

\paragraph{Optimisations leveraging linearity.}
We only briefly mentioned how linearity can inform optimisations to produce
more performant programs. We leave exploring optimisations unblocked by
preserving linearity in the intermediate language with Linear Core as future
work. Linearity influenced optimising transformations have been also discussed
by Linear Haskell~\cite{cite:linearhaskell} and
in~\cite{cite:let-floating,peytonjones1997a}. An obvious candidate is
\emph{inlining}, which is applied based on heuristics from information provided
by the \emph{cardinality analysis} pass that counts occurrences of bound
variables.  Linearity can be used to non-heuristically inform
the inliner~\cite{cite:linearhaskell}.
% Additionally, we argue that in Linear Core accepting more programs as linear
% there are more chances to use linearity, in contrast to a linear type system
% which does not account for lazy evaluation and thus rejects more programs.

% \paragraph{Usage environment resource inference.}
% In Section~\ref{sec:linearity-semantically}, we discussed how there is not an obvious way to determine the set of linear
% resources used by a group of (mutually) recursive bindings, .
% though the resources used by each binder are the solution to a set determined
% by the recursive bindings group.  In Section~\ref{sec:main:linear-core}, we
% further likened the challenge of determining usage environments for a recursive
% group of bindings to a unification problem as that solved by the Hindley-Milner
% type inference algorithm~\cite{DBLP:conf/popl/DamasM82} based on generating and solving
% constraints. Even though these are useful observations, our implementation of
% Linear Core uses a naive algorithm to determine the usage environments, thereby
% leaving as future work the design of a principled algorithm to determine the
% usage environments of recursive group of bindings.

### 2025-10-07 popl26-accepted-before-rebuttal.tex

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
Linear Core is suitable as the intermediate language of an optimising compiler
for a linear and lazy language such as Haskell Core, in that optimising
transformations in Linear Core preserve types \emph{and} linearity, since Linear
Core understands semantic linearity in the presence of laziness, unlike
Core's current type system under which optimisations currently violate
linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
A linearly typed Core that preserves linearity throughout the optimisation
pipeline of GHC both validates the correctness of the compiler, which is
already achieved to a great extent by preserving (non-linear) types, and
informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{[Link removed
due to
anonymization]} %\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218} ,
we leave this as future work.

\paragraph{Generalizing Linear Core to Haskell.}
Linear types, despite their compile-time correctness guarantees regarding
resource management, impose a burden on programmers in being a restrictive
typing discipline (witnessed, e.g., by Linear
Constraints~\cite{cite:linearconstraints}). Linear Core eases the restrictions
of linear typing by being more flexible in understanding linearity for lazy
languages. In this sense, it is an avenue of future
work to apply the ideas from this work to the surface Haskell language.

% \section*{Acknowledgements}

% We thank Arnaud Spiwack for valuable and insightful discussions.\todo{Do these have to be anonymized?}

% \begin{itemize}
% 
% \item Linear(X), a linear type system defined by the underlying definition of
% evaluation (which in turn implies how consuming a resource is defined)
% 
% \item Implementation in Core
% 
% \item Generalization to source level language, being more permissive in the
% handling of resources imposes less burden on the programmer
% 
% \item It's harder to typecheck linearity like this in the source level because
% of the interaction with other source features, but seems feasible and an
% improvement to the usability of linear types. It allows more lazy functional
% programming idioms with linear types (also because laziness and strictness is less well defined as in Core, bc opts)
% 
% \item Beautiful inference algorithm for recursive usage environments -- insight
% that looks like inference for recursive function principle types, but haven't
% figured it out -- connection to type inference / hindley milner
% 
% \item We kind of ignore the multiplicity semiring. We should discuss
% how we don't do ring operations ... but that's kind of wrong.
% 
% % \item We know the case binder to ALWAYS be in WHNF, perhaps there could
% % be some annotation on the case binder s.t. we know nothing happens when we
% % scrutinize it as a single variable
% 
% \item Mechanizing the system and metatheory
% 
% \end{itemize}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Conclusion
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% We give up on this section and instead describe "syntax-directedness" in the implementation chapter
%
%
% \section{Syntax Directed System}
% 
% \todo[inline]{In the other system we assume that the recursive lets are strongly connected, i.e. the expressions always}
% 
% \input{language-v4/SyntaxDirectedSystem}
% 
% \subsection{Inferring usage environments}
% 
% \todo[inline]{The difference between this and the previous section is a bit blurry}
% 
% \todo[inline]{There's one more concern: usage environments aren't readily
% available, especially in recursive lets. We must perform inference of usage
% environments before we can typecheck using them. This is how:}
% 
% \todo[inline]{Rather, we define a syntax directed type system that infers usage environments while checking...}
%

% \section{Conclusion}

% Optimising compilers with a typed and lazy intermediate language with linear
% types (of which GHC is the prime example) leverage laziness to heavily
% transform and rewrite programs into simpler forms.
% %
% However, these optimising transformations push the interaction between
% linearity and laziness to the limits where linearity can no longer be seen
% syntactically, despite being maintained semantically, in the sense that linear
% resources are still used exactly once when the optimised program is run.

% In this work we explored linearity in the presence of laziness by example
% through the interactions of linear types with lazy (recursive) let bindings and
% case expressions that evaluate their scrutinee to Weak Head Normal Form. Most
% example programs were linear semantically, but not syntactically.
% %
% We developed a linear type system, Linear Core, for an intermediate language
% akin to GHC Core, with laziness and linearity. In contrast to GHC Core's type
% system, or any other linear type system (to the best of our knowledge), our
% type system understands semantic linearity, and can thus correctly type a wider
% range of linear programs, as those explored in the semantic linearity examples.
% %
% Crucially, we proved soundness of the type system, and proved multiple
% optimising transformations preserve linearity, despite most violating linearity
% in other linear type systems. Additionally, we implemented Linear Core as a GHC
% plugin to further explore its suitability in the intermediate language of an
% optimising compiler.

% Concluding, Linear Core is a suitable type system for linear, lazy,
% intermediate languages of optimising compilers such as GHC, as it understands
% linearity in the presence of laziness s.t. optimisations preserve types and
% linearity, and further unblocks optimisations influenced by linearity, e.g.
% inlining and call-by-name $\beta$-reduction for applications of (semantically)
% linear functions.

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\section{Natural and Instrumented Operational Semantics}\label{app:opsem}

We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
equipped with an environment which assigns expressions to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation that makes all sharing explicit
through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.
\begin{figure}
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                       \qquad\mbox{with
                                       $\ov{\Gamma;\Delta_i\vdash
                                       e_i :
                                       \sigma_i}$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
    \]
    \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
  \end{figure}

The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
  \]
  \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
\end{figure}

Our instrumented, linearity-aware semantics (akin to that
of~\cite{cite:linearhaskell})
is defined in Figure~\ref{fig:instrsemapp}.
In this semantics, linear variables are erased from
the runtime environment, ensuring that any term violating linearity
produces a stuck state. The semantics is defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions) for variables occurring in $\Sigma$;
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$, which are expressions in which variables in $\Theta$ are
used.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
        \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
    \quad
            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
      \D'' = \D' \setminus (\Theta \setminus \Theta')
      }
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta'}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau , \Sigma
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~\ov{x_i}) \Downarrow
    (\Theta \mid K~\ov{x_i}) : T , \Sigma}
    \\[1.5em]

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]
